### 2013-05-18 draft.tex
fixed the stratigraphic trends section of the discussion - says more about d18O, less about diagenesis

In addition to being highly sensitive to diagenetic alteration, carbonate clumped isotopes are also diagnostic of diagenetic regimes, ultimately allowing identification of the most pristine endmember of a sample set. Clumped isotopes can distinguish between diagenesis in the presence of low and high water-to-rock ratios (rock-buffered and water-buffered diagenesis, respectively) and diagenesis in the presence of meteoric water. As carbonates are buried deeper and the temperature increases, the equilibrium oxygen isotope fractionation between water and carbonate grows smaller, preferentially driving $^{18}$O into the water. During rock-buffered diagenesis, this causes the \deltao O of the water to increase, while the \deltao O of the altered carbonate is not affected. In a water-buffered diagenetic regime, the \deltao O of the water does not change, so increasing temperature instead causes the \deltao O of the altered carbonate to decrease. In constrast, diagenesis in the presence of meteoric water usually occurs when the sample is near the Earth's surface, so the clumped isotope temperature of the altered carbonate is cool, but the \deltao O of the water is diagnostically low. The ability of clumped isotopes to constrain the conditions to which fossils have been exposed after burial allows identification of the best-preserved samples from within a set of samples that traditionally would all be considered pristine \citep{Eiler2011}. We demonstrate here that this distinction has important implications for the estimation of the temperature and \deltao O of oceans in the past.

@@

All $\Delta_{47}$ values are projected into the absolute reference frame \citep{Dennis2011}. When possible, raw $\Delta_{47-[\text{SGvsWG}]}$ values are converted to the absolute reference frame using an empirical transfer function (ETF) constructed from the heated and equilibrated  gases measured during the same week as the samples. Heated gases are aliquots of CO$_2$ that were sealed in quartz tubes, heated at 1000\degrees C for three hours, and then cooled rapidly to room temperature. Equilibrated gases are aliquots of CO$_2$ that were equilibrated with water at 25\degrees C for at least 24 hours. Heated gases were made from two separate reservoirs of CO$_2$ with different \deltao O compositions, and equilibrated gases were equilibrated with either of two separate waters, also differing in their \deltao O compositions. Heated and equilibrated gases were cleaned following the same procedure as the samples. The ETF was constructed and then used to convert raw $\Delta_{47-[\text{SGvsWG}]}$ values following \cite{Dennis2011}. Finally, the resulting $\Delta_{47-\text{RF}}$ was corrected for the acid extraction at 90\degrees C by adding 0.092$\permil$ \citep{Henkes2013}.

@@

The \deltac C values of fossil material, micrite and cement are highly consistent within beds (Figure \ref{section}). We resolve a 3$\permil$ positive excursion in \deltac C, beginning just below the boundary between the Lower and Upper Visby Beds, and continuing for at least 50 meters. The sample set does not capture the relaxation of this excursion, but our samples from the Slite Formation, beginning at 76 meters, record a return to lower \deltac C values of approximately -0.5$\permil$. The \deltac C trend through the section is comparable to the brachiopod \deltac C measurements from \cite{Bickert1997} and \cite{Munnecke2003}.

The \deltao O values of fossil material is consistent within beds, but micrite and cements tend to have lower \deltao O values than fossil material, often several permil lower than fossils in the same bed (Figure \ref{section}). We were not able to resolve a \deltao O excursion from our sample set, but the range in \deltao O values of fossil material is comparable to the range observed by \cite{Bickert1997} and \cite{Munnecke2003}.

The clumped isotope temperature of samples is highly variable within beds, often covering a range of 10\degrees C within a single bed (Figure \ref{section}). The lowest temperature sample, a brachiopod, and the highest temperature sample, a rugose coral, are both from the Slite Formation, deposited well after the hypothesized climate phenomenon associated with the isotope excursions in the Lower and Upper Visby Beds. Due to the scatter in temperature, we were not able to resolve a change in temperature with time.

The large scatter in temperatures also suggests variable preservation quality among the samples. Even brachiopods, which are commonly thought to be resistant to diagenetic alteration, have a range in temperature of more than 15\degrees C. Diagenetic alteration can occur at elevated temperatures deep underground, which can increase the clumped isotope temperature of altered samples, but alteration can also occur near the Earth's surface, in the presence of cool, likely meteoric water. Consequently, low-temperature samples cannot be assumed to be more pristine than high-temperature samples. However, alteration in the presence of cool, meteoric water will deplete the \deltao O of cabonate, because meteoric water usually has a very light \deltao O. Most of our fossil samples are very similar in \deltao O, making alteration by meteoric water unlikely. This leads to a hypothesis that the lowest temperature samples are the most pristine, while the higher temperature samples likely represent diagenetic alteration during burial at high temperature. We tested this hypothesis using independent methods to assess the preservation quality of representative samples.

>>>>

In addition to being highly sensitive to diagenetic alteration, carbonate clumped isotopes are also diagnostic of diagenetic regimes, ultimately allowing identification of the most pristine endmember of a sample set. Clumped isotopes can distinguish between diagenesis in the presence of low and high water-to-rock ratios (rock-buffered and water-buffered diagenesis, respectively) and diagenesis in the presence of meteoric water. As carbonates are buried deeper and the temperature increases, the equilibrium oxygen isotope fractionation between water and carbonate grows smaller, preferentially driving $^{18}$O into the water. During rock-buffered diagenesis, this causes the \deltao O of the water to increase, while the \deltao O of the altered carbonate is not affected. In a water-buffered diagenetic regime, the \deltao O of the water does not change, so increasing temperature instead causes the \deltao O of the altered carbonate to decrease. In constrast, diagenesis in the presence of meteoric water usually occurs when the sample is near the Earth's surface, so the clumped isotope temperature of the altered carbonate is cool, but the \deltao O of the carbonate and the water is diagnostically low. The ability of clumped isotopes to constrain the conditions to which fossils have been exposed after burial allows identification of the best-preserved samples from within a set of samples that traditionally would all be considered pristine \citep{Eiler2011}. We demonstrate here that this distinction has important implications for the estimation of the temperature and \deltao O of oceans in the past.

@@

All $\Delta_{47}$ values are projected into the absolute reference frame \citep{Dennis2011}. When possible, raw $\Delta_{47-[\text{SGvsWG}]}$ values are converted to the absolute reference frame using an empirical transfer function (ETF) constructed from the heated and equilibrated  gases measured during the same week as the samples. Heated gases are aliquots of CO$_2$ that were sealed in quartz tubes, heated at 1000\degrees C for three hours, and then cooled rapidly to room temperature. Equilibrated gases are aliquots of CO$_2$ that were equilibrated with water at 25\degrees C for at least 24 hours. Heated gases were made from two separate reservoirs of CO$_2$ with different \deltao O compositions, and equilibrated gases were equilibrated with either of two separate waters, also differing in their \deltao O compositions. Heated and equilibrated gases were cleaned following the same procedure as the samples. The ETF was constructed and then used to convert raw $\Delta_{47-[\text{SGvsWG}]}$ values to the absolute reference frame following \cite{Dennis2011}. Finally, the resulting $\Delta_{47-\text{RF}}$ was corrected for the acid extraction at 90\degrees C by adding 0.092$\permil$ \citep{Henkes2013}.

@@

The \deltac C values of fossil material, micrite and cement are highly consistent within beds (Figure \ref{section}). We resolve a 3$\permil$ positive excursion in \deltac C, beginning just below the boundary between the Lower and Upper Visby Beds, and continuing for at least 50 meters. The sample set does not capture the relaxation of this excursion, but our samples from the Slite Formation, beginning at 76 meters, record a return to lower \deltac C$_{\text{PDB}}$ values of approximately -0.5$\permil$. The \deltac C trend through the section is comparable to the brachiopod \deltac C measurements from \cite{Bickert1997} and \cite{Munnecke2003}.

The \deltao O values of fossil material are consistent within beds, but micrite and cements tend to have lower \deltao O values than fossil material, often several permil lower than fossils in the same bed (Figure \ref{section}). We are not able to resolve an excursion in \deltao O, although this may simply be because our dataset contains fewer samples than that of \cite{Munnecke2003}. The average \deltao O$_{\text{PDB}}$ of the fossils remains fairly constant at -5.5$\permil$ through the Lower and Upper Visby Beds, making our Lower Visby samples consistent with those of \cite{Munnecke2003} but our Upper Visby samples more depleted in $^{18}$O. Higher in the section, the \deltao O of the fossils in this study appear consistent with values reported by \cite{Bickert1997}.

The clumped isotope temperature of samples is highly variable within beds, often covering a range of 10\degrees C within a single bed (Figure \ref{section}). The lowest temperature sample, a brachiopod, and the highest temperature sample, a rugose coral, are both from the Slite Formation, deposited well after the hypothesized climate phenomenon associated with the isotopic excursions in the Lower and Upper Visby Beds. Due to the scatter in temperature, we are not able to resolve a change in temperature with time.

The large scatter in temperatures also suggests variable preservation quality among the samples. Even brachiopods, which are commonly thought to be resistant to diagenetic alteration, have a range in temperature of more than 15\degrees C. All of the fossil samples have fairly high values of \deltao O, making alteration by meteoric water unlikely. This leads to a hypothesis that the lowest temperature samples are the most pristine, while the higher temperature samples likely represent diagenetic alteration during burial at high temperature. We tested this hypothesis using independent methods to assess the preservation quality of representative samples.
